Adaptive evolution with aneuploidy and mutation

### 2022-03-21 ms/ms.tex
revised a bit

Transitions between the genotypes occur as follows~(\Cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \Cref{table:single-locus}.

%%% Table: single-locus model
\begin{table}[h]
\centering
\caption{\textbf{Single-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
\end{tabular}
\label{table:single-locus}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \Cref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

@@

% YR: this whole section doesn't really make sense. We know there was aneuploidy. If we assume delta=0 that means that we think this aneuploidy was not relevant to the adaptation process. But then why do we take generations 1700 and 2300 to be the time for 2n* fixation? These times were not observed for fixation of a mutant but rather for elimination of aneuploidy. If we disregard aneuploidy, why do we still consider these times?
%For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), 
%the likelihood is similarly approximated by
%\begin{equation}\begin{aligned}
%\label{eq:heatstress-noaneuploidy-likelihood}
%\likelihood_{!}(\theta) = &\ 
%	1 - 
%	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
%&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
%&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
%\;.
%\end{aligned}\end{equation}

@@

We also plot posterior predictions: for each model we execute $10,000$ simulations using the MAP parameter estimates and plot the distributions of time to fixation of \eumt, one of key properties of the model likelihood. This visualizes the fit of each model to the data. 
Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models are interchangeable. We validate this by comparing HDI (highest density interval) of posterior distributions.

\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs. 3C, 4A, and S2]{Yona2012}.
We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed. Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix. We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\Cref{fig:growth-curves}-D). In lieu of a more suitable prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results.

Also we tried extended informative prior that use additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$ ~(\Cref{fig:growth-curves}L). The same prior was used for $w_{2n*}/w_{2n+1*}$.

Also we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In both cases the inference algorithm failed to converge.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets.
%% estimates generated by MAP-and-hdi.ipynb
The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$, agrees with previous estimates. % TODO refs (9.7*10-5 in Zhu 2014, so it's lower). YR: we need more than one ref; also what is the units of Zho 2014?
The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated fitness values are $w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.025\ [1.024-1.026]$,
$w_{2n^*}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$. 
Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:single-locus}).

@@

% commented out, see remark above when describing likelihood of delta=0.
%Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). %% generated with MAP-and-hdi.ipynb
%The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40. 
%The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
%This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations. Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).

@@

% generated with alt-prior.ipynb
At the end we inferred parameters for \emph{extended informative prior} model (see \emph{Prior distributions}) with different priors for $w_{2n+1^*}$ and $w_{2n*}$. The estimated mutation rate was much lower compared to the basic model, $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
$w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.032\ [1.03-1.035]$,
$w_{2n^*}=1.033\ [1.031-1.036]$. 
The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}).

% Fig single-locus model with and without aneuploidy
%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
\begin{figure}[p]
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
      \label{fig:fit}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
      \label{fig:fit-cumulative}
  \end{subfigure}
  \caption{
    \textbf{Single-locus model fit with and without aneuploidy.}
    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450. Thus, the blue distribution is a better fit compared to the green, and the yellow histogram has a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
    \textbf{(B)}
	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A. The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
     }
  \label{fig:2n*-fixation}
\end{figure}

\begin{figure}[p]
  \centering
\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
  \caption{
  \textbf{Posterior genotype frequency dynamics for the single-locus model.} 
  The posterior prediction for the frequencies of the four genotypes over time. Each of the 20 curves is the average of 10,000 simulations of the single-locus model using parameters drawn from the posterior distribution.
  }
  \label{fig:ppc-plot}
\end{figure}

% Fig tau plots
%% generated by diff-tau-plots.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-a.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-b.pdf}      
  \end{subfigure}
   \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-c.pdf}      
  \end{subfigure}
   \caption{
    \textbf{Genotype fixations for single-locus models with different $\tau$.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells. We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is fixed (y-axis) at each generation (x-axis). Note that \anmt\ did not fix. We can see that $\tau=100$ can be distinguished if time to loss of \euwt\ is known (panel~A) or if time to fixation or loss of \anwt\ is known (panel~B). It is harder to distinguish between $1\leq \tau \leq10$.
  }
  \label{fig:tau-plots}
\end{figure}

@@

Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) and analyse how it is affected by the model parameters.
As a stochastic model, there are two events that can occur on the \euwt\ genotype background. The first is a mutation that produces a \eumt\ genotype. The second is a missegregation event that produces a \anwt\ genotype. The probabilities with which these events occur is significantly different. We are interested in the average waiting time until one of these two genotype appears and then leads to adaptation, i.e., appearance and fixation of the \eumt; in the second case, appearance of \anwt\ will be followed by a mutation to \anmt\ and loss of aneuploidy to \eumt.

We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt\ are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
\begin{subequations}
\begin{align}
p_{2n^*} &= 1-(1-2\mu \Delta w_{2n^*})^N \;, \\
p_{2n+1} &= 1-(1-2\delta\Delta w_{2n+1})^N \;,
\end{align}
\end{subequations}
where $\Delta w_i$ denotes the relative difference between the fitness $w_i$ of genotype $i$ and the population mean fitness, $\bar{w}$.

However, we have two competing events: if one event occurs (e.g., mutation), then the other event (e.g., aneuploidy) is unlikely to occur. Thus, we need to find the minimum of these waiting times. The minimum of two geometric random variables with rates $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with rate $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.

% TODO MP: this seems like repetition of things written above, right? should we remove this? or if there is something important here, merge it to previous paragraphs?
%We are interested in the expected time to fixation of \eumt, and the trajectory leading to this fixation depends on the event that occurs first. If \eumt\ appears first and avoids extinction by drift, then it will fix. If \anwt\ appears first and avoids extinction, then several other events must occur for fixation of \eumt. Thus, in this aneuploid trajectory, we need to approximate the expected waiting time to reach \anmt\ and finally \eumt.

Given a mutation occurs first, % TODO MP: I added this sentence, is that right?
the expected adaptation time via the mutation trajectory is 
\begin{equation}\label{eq:T_mut}
T_{M}=\frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + T_{2n \to 2n^*} \;,
\end{equation}
where $T_{i \to j}$ is the waiting time for fixation of genotype $i$ in a population of genotype $j$, given it starts with a single copy and does not go to extinction by drift.
Given that aneuploidy occurs first, % TODO MP: I added this sentence, is that right?
the expected adaptation time via the aneuploidy trajectory is approximated by
\begin{equation}\label{eq:T_aneu} 
% TODO MP: p_{2n+1^*} is not defined, and why is there 1/p_{2n^*} here? shouldn't it have the waiting time for 2n+1* to appear on 2n+1 and then waiting time for 2n* to appear on 2n+1*? something is wrong here.
 T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{2n+1^*}} + \frac{1}{p_{2n^*}} + T_{2n \to 2n+1} + T_{2n+1 \to 2n+1^*} + T_{2n+1^* \to 2n^*} \;.
 \end{equation} % TODO MP do we need the tilde ?
Finally, the expected adaptation time is 
 \begin{equation}\label{eq:T_total}
 T = p_A T_{A}+(1-p_A) T_{M} \;,
 \end{equation}
 where $p_A$ is he probability that aneuploidy will be the first event to occur,
\begin{equation}
p_A = \frac{p_{2n+1}}{p_{2n^*}+p_{2n+1}} \;.
\end{equation}

We assume that the waiting time until a new genotype occurs is much longer than the waiting time for its fixation in the population. Nevertheless, these fixation times must be included in a rigorous analysis. There are two ways to compute the fixation times and they differ in the assumption of the underlying process.
A deterministic-process approach is given by the classical bi-allelic single-locus equation.
It approximates the expected time for genotype $i$ to reach frequency $x$ when starting from frequency $1-x$ in a population of genotype $j$,
 % TODO MP ref for equation? Crow and Kimura? YR: yes
\begin{equation} \label{eq:T_fix_x}
 T_{i \to j} \approx 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
\end{equation} 
A stochastic-process approach for approximating $T_{i \to j}$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}.

% TODO MP - the following remark should be removed. Instead, wherever you use one of these assumptions in the above text, you should explicitly say so (rather then say the assumptions are implicit). Then you can discuss here how these assumptions may not be met when the parameters are so and so, and that you will point it out in the results or something like that.
 %\begin{remark} There are several implicit assumptions that the equations hold like this.
 %\begin{enumerate}
 %\item The two processes fixation and occurrence of a new type are sequential and do not interfere with each other.
 %\item There is no clonal interference.
 %\item The fixation process starts from exactly one individual. 
 %\end{enumerate}
 %There are certainly combinations of the mutation rate, the aneuploidy rate, the selection strength and the population size for which at least one of these assumptions is violated. Thus, \Cref{eq:T_total} may not yield a good approximation to the actual time to the adapted state.
 %\end{remark}

First, to validate the approximation in \Cref{eq:T_total}, we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single}) without aneuploidy, i.e., $\delta=0$ (\Cref{fig:only mut}), and with the other parameters as inferred above. The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells. In that case, the expected number of new mutants per generation is roughly 0.5, and  we can expect that in many generations the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which violates our assumption that an adaptation event starts from a single mutant cell.

 \begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{../figures/figgrid_noaneu.pdf}
 \caption{\textbf{Expected adaptation time without aneuploidy.}
  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (\textbf{A}) or as a function of the mutation rate, $\mu$ (\textbf{B}). Mean and 95\% CI from simulations in blue. Analytic approximation (\Cref{eq:T_total}) in orange. Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$. In panel A, $\mu=9.6\cdot 10^{-8}$. In panel B, $N=6\cdot 10^6$.
  }
  %TODO MP 
  % which approximation do you use here for T_i->j ?
  % The error bars a quite large, can you make them smaller by running more simulations?
  % --> or by making the y axis in log scale!!
 \label{fig:only mut}
 \end{figure}

 \begin{figure}[h!]
 %\centering
 \includegraphics[width=1\textwidth]{../figures/figgrid_nodirectmut_ylog_uncorr0.jpg}
 \caption{\textbf{Expected adaptation time without mutations from $\euwt$ to $\eumt$.}
  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (\textbf{A}) and as a function of the mutation rate, $\mu$ (\textbf{B}) and as a function of the aneuploidy rate, $\delta$ ({C}). Blue error bars for mean and 95\% CI from simulations. Green line for the analytic approximation (\Cref{eq:T_total}). % TODO why green? in previous figure it was orange.
  Red curve for the approximation with $T_{\euwt \to \anwt}=T_{\anwt \to \anmt}=0$. 
  Orange curve for the approximation with $x=0.017$ in \Cref{eq:T_fix_x} for the fixation time of $\eumt$; we found this gives a good fit in panel B.
  Here, $w_\anwt=1.021,\, w_\anmt=1.025$ and $w_\eumt=1.028$, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$ unless parameter is varied on the x-axis.
  }
 \label{fig:no direct mut}
 \end{figure}

>>>>

Transitions between the genotypes occur as follows~(\Cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \Cref{table:fitness}.

%%% Table: fitness
\begin{table}[h]
\centering
\caption{\textbf{Fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
\end{tabular}
\label{table:fitness}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \Cref{table:fitness} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

@@

For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), we disregard the increased expression in chromosome III and the growth advantage measured in generation 450, and focus on the growth advantage measured in later generations, presumably due to a beneficial mutation. 
Therefore, the likelihood is approximated by
\begin{equation}\begin{aligned}
\label{eq:heatstress-noaneuploidy-likelihood}
\likelihood_{!}(\theta) = &\ 
	1 - 
	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
\;.
\end{aligned}\end{equation}

@@

We also plot posterior predictions: for each model we execute $10,000$ simulations using the MAP parameter estimates and plot the distributions of time to fixation of \eumt, one of key properties of the model likelihood. These plots visualize the fit of each model to the data. 
Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models are interchangeable. We validate this by comparing HDI (highest density interval) of posterior distributions.

\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs. 3C, 4A, and S2]{Yona2012}.
We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed. Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.

We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$~(\Cref{fig:growth-curves}-D, assuming $w_{2n}=1$). In lieu of a more suitable prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results. 
We also examined an extended informative prior that uses additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$~(\Cref{fig:growth-curves}L). The same prior was used for $w_{2n*}/w_{2n+1*}$.

As a control, we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In these cases the inference algorithm failed to converge.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets.
%% estimates generated by MAP-and-hdi.ipynb
The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$, agrees with previous estimates. % TODO refs (9.7*10-5 in Zhu 2014, so it's lower). YR: we need more than one ref; also what is the units of Zho 2014?
The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated fitness values are $w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.025\ [1.024-1.026]$,
$w_{2n^*}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$. 
Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:fitness}).

@@

% commented out, see remark above when describing likelihood of delta=0.
Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). %% generated with MAP-and-hdi.ipynb
The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40. 
The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations. Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).

@@

% generated with alt-prior.ipynb
Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with different priors for $w_{2n+1^*}$ and $w_{2n*}$. The estimated mutation rate was much lower compared to the basic prior, with $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
$w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.032\ [1.03-1.035]$,
$w_{2n^*}=1.033\ [1.031-1.036]$. 
The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}).

% Fig model with and without aneuploidy
%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
\begin{figure}[p]
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
      \label{fig:fit}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
      \label{fig:fit-cumulative}
  \end{subfigure}
  \caption{
    \textbf{Model fit with and without aneuploidy.}
    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450. Thus, the blue distribution is a better fit compared to the green, and the yellow histogram has a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
    \textbf{(B)}
	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A. The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
     }
     % TODO should we remove this figure altoghther? change it? because we removed no-aneuploidy model...? should we merge this with the next figure?
  \label{fig:2n*-fixation}
\end{figure}

\begin{figure}[p]
  \centering
\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
  \caption{
  \textbf{Posterior genotype frequency dynamics for the model.} 
  The posterior prediction for the frequencies of the four genotypes over time. Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
  }
  \label{fig:ppc-plot}
\end{figure}

% Fig tau plots
%% generated by diff-tau-plots.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-a.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-b.pdf}      
  \end{subfigure}
   \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-c.pdf}      
  \end{subfigure}
   \caption{
    \textbf{Genotype fixations for models with different $\tau$.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells. We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is fixed (y-axis) at each generation (x-axis). Note that \anmt\ did not fix. We can see that $\tau=100$ can be distinguished if time to loss of \euwt\ is known (panel~A) or if time to fixation or loss of \anwt\ is known (panel~B). It is harder to distinguish between $1\leq \tau \leq10$.
  }
  \label{fig:tau-plots}
\end{figure}

@@
